% =========================
% External / Related Works (one-shot section for KI; BibLaTeX)
% Include once in KI, anywhere appropriate, e.g., before \printbibliography.
% Requires \addbibresource{external_related_works.bib} in the preamble.
% =========================
\section{External / Related Works}

\paragraph{Shell and fixed-point logics.}
We situate our \emph{shell} invariants alongside classic modal $\mu$-calculus and model checking texts
(\parencite{Kozen1983,BradfieldStirling2007,BlackburnDeRijkeVenema2001,BaierKatoen2008,ClarkeGrumbergPeled1999}).

\paragraph{Categorical probability and decision kernels.}
Our kernel semantics follows Markov/categorical probability
(\parencite{Giry1982,Fritz2020,Jacobs2011,Staton2017}).

\paragraph{Drift, ergodicity, and mixing.}
Stability bounds use classical drift/minorization and mixing estimates
(\parencite{MeynTweedie2009,LevinPeresWilmer2009,Norris1998,Hajek1982}).

\paragraph{Aggregation and coarse-graining.}
The quotient/aggregation moves align with lumpability and metastability
(\parencite{KemenySnell1976,Seneta2006,DeuflhardWeber2005,LiWalshLittman2006}).

\paragraph{Spectral perturbation and continuity.}
Continuity under constraint shifts leverages operator and matrix perturbation theory
(\parencite{Kato1995,StewartSun1990}).

\paragraph{Large deviations and tilted operators.}
Concentration and scaled cumulants connect to Donsker--Varadhan theory and LDP
(\parencite{DonskerVaradhan1975,DemboZeitouni1998,Ellis1985}).

\paragraph{Information transport and Fisher geometry.}
Identifiability, sufficiency, and curvature follow information-theoretic and information-geometric foundations
(\parencite{Blackwell1953,LehmannCasella1998,CoverThomas2006,AmariNagaoka2000}).

\paragraph{Temperament priors and control.}
Policy priors and free-energy objectives echo CMDP/KL-control and PAC-Bayes calibration
(\parencite{Puterman1994,Altman1999,Kappen2005,Todorov2009,McAllester1999,Catoni2007}).

\paragraph{Flow centrality.}
Action-flow betweenness connects to centrality indices and electrical-network analogies
(\parencite{Freeman1977,Katz1953,Newman2005,DoyleSnell1984}).

\paragraph{Protocol layer.}
For streaming detection and multiplicity control we reference Page, Basseville--Nikiforov, Benjamini--Hochberg,
and modern online-FDR developments
(\parencite{Page1954,BassevilleNikiforov1993,BenjaminiHochberg1995,JavanmardMontanari2018,RamdasEtAl2018}).